Extreme value analysis for the sample autocovariance matrices of heavy-tailed multivariate time series

### 2017-08-21 papers/DavisMikoschHeinyXie/PaperBody.tex
Revised the phd thesis.

To illustrate the theory we consider a simple moving average example
taken from Davis et al.~\cite{davis:mikosch:pfaffel:2015}.
\begin{example} \rm
\begin{figure}[htb!]
  \centering
  \subfigure[iid data]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps_iid.pdf}
  }
  \subfigure[data from model \eqref{ex:4.6}]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps.pdf}
  }
  \caption{Distribution function of $(\lambda_{(1)} - \lambda_{(2)})/\lambda_{(1)}$ for iid data (left) and
    data generated from the model \eqref{ex:4.6} (right). In each graph we compare the empirical \ds\ \fct\
(dotted line, based on
1000 simulations of $200 \times 1000$ matrices with $Z$-\ds\ \eqref{eq:distrsim}) with the theoretical curve (solid line).}
  \label{fig:ProbMass}
\end{figure}
Assume that $\alpha\in (0,2)$ and
\begin{equation}\label{ex:4.6}
X_{it}= Z_{it}+ Z_{i,t-1}-2 (Z_{i-1,t}- Z_{i-1,t-1})\,,\quad i,t\in\Z\,.
\end{equation}
In this case, the non-zero entries of $\H$ are
\begin{equation*}
h_{00}=1, h_{01}=1,h_{10}=-2 \quad \mbox{ and }\quad h_{11}=2.
\end{equation*}
Hence $\M=\H\H'$ has the positive eigenvalues
$v_1=8$ and $v_2=2$. The limit point process in \eqref{eq:Davispp} is
\begin{equation*}
N=\sum_{i=1}^\infty \vep_{8\Gamma_i^{-2/\alpha}}+ \sum_{i=1}^\infty \vep_{2\Gamma_i^{-2/\alpha}}\,,
\end{equation*}
so that
\begin{equation*}
a_{np}^{-2}\,\big(\la_{(1)},\la_{(2)}\big) \cid
\big(8\Gamma_{1}^{-2/\alpha}, 2\Gamma_1^{-2/\alpha}\vee 8
\Gamma_2^{-2/\alpha}\big)\,.
\end{equation*}
Using the fact that $U=\Gamma_1/\Gamma_2$ has a uniform \ds\ on $(0,1)$ we calculate
\begin{equation*}
\P(2\Gamma_1^{-2/\alpha}>8\Gamma_2^{-2/\alpha})= \P(\Gamma_1/\Gamma_2<2^{-\alpha})= 2^{-\alpha} \in (1/4,1).
\end{equation*}
In particular, we have for the normalized spectral gap
\begin{equation*}
a_{np}^{-2} \big(\la_{(1)}-\la_{(2)}\big)\cid
6 \,\Gamma_1^{-2/\alpha} \1_{\{\Gamma_1 4^{\alpha/2 }<\Gamma_2\}}+
8\,
\big(\Gamma_1^{-2/\alpha}-\Gamma_2^{-2/\alpha}\big)\1_{\{\Gamma_14^{\alpha/2
  }> \Gamma_2\}}
\end{equation*}
and for the self-normalized spectral gap (see also Example~\ref{exam:spectralgap} for a detailed analysis)
\begin{equation*}
\begin{split}
\dfrac{\la_{(1)}-\la_{(2)}}{\la_{(1)}}&\cid
\dfrac{6}{8} \, \1_{\{\Gamma_1 2^{\alpha }<\Gamma_2\}}+
\big(1-(\Gamma_1/\Gamma_2)^{2/\alpha}\big)\1_{\{\Gamma_12^{\alpha
  }> \Gamma_2\}}\\
& =\dfrac{3}{4}\,  \1_{\{U 2^{\alpha }<1\}}+
\big(1-U^{2/\alpha}\big)\1_{\{U2^{\alpha
  }> 1\}}=Y\,.
\end{split}
\end{equation*}
The limit  distribution of the spectral gap
has an atom at $3/4$ with probability $2^{-\alpha}$, i.e.,~$\P(Y=3/4)=2^{-\alpha}$, and
\begin{equation*}
\P(Y\le x)= 1- (1-x)^{\alpha/2},\qquad x\in (0,3/4).
\end{equation*}
In the iid case the limit distribution of the self-normalized spectral gap has distribution function $F(x) = 1- (1-x)^{\alpha/2}$ for $x\in [0,1]$. This means that the atom disappears if the entries are iid. Figure~\ref{fig:ProbMass} compares the distribution function of $Y$
with $F$ for $\alpha=0.6$; the atom at $3/4$ is clearly visible.

@@

8\,\Gamma_1^{-2/\alpha},\frac{1}{4}\,\1_{\{U<2^{-\alpha}\}}+U^{2/\alpha}\,\1_{\{U\ge

@@

uniform \ds , $2^{-\alpha}\, \1_{\{U<2^{-\alpha}\}}+U\, \1_{\{U\ge 2^{-\alpha} \}}$.

@@

\frac{v_2}{v_1}\,\1_{\{U<(v_2/v_1)^{\alpha/2}\}}+U^{2/\alpha}\,\1_{\{U\ge (v_2/v_1)^{\alpha/2}\}}\,.

@@

\dfrac{\la_{(1)}-\la_{(2)}}{\la_{(1)}} \cid \dfrac{v_1-v_2}{v_1}\,\1_{\{U<(v_2/v_1)^{\alpha/2}\}}+(1-U^{2/\alpha})\,\1_{\{U\ge (v_2/v_1)^{\alpha/2}\}}\,.

@@

\subsection{S\&P 500 data}\label{sec:sp500}
We conduct a short analysis of the largest eigenvalues of the univariate log-return time series which compose
the S\&P 500 stock index; see Section~\ref{subsec:1.2} for a description of the data.
Although there is strong empirical evidence that these univariate series have power-law tails
(see Figure~\ref{fig:SP500_tail_indices}) we  do not expect that they
have the same tail index. One way to proceed would be to ignore this fact because the tail indices are
in a close range and the differences are due to large sampling errors for estimating such quantities.  One could also collect \ts\ with similar
tail indices in the same group.
In this case, the dimension $p$ decreases. This grouping would be a rather arbitrary classification method.
We have chosen a third way: to use rank transforms.
This approach has its merits because it aims at standardizing the tails but it also has a major disadvantage: one destroys
the covariance structure underlying the data.
\par
Given a $p\times n$ matrix $(R_{it})_{i = 1, \cdots, p;t=1,\cdots, n}$,
we construct a matrix $\X$ via the rank transforms
\begin{equation*}
  X_{it} = -\Big[
    \log \Big(\frac{1}{n+1} \sum_{\tau=1}^n \1_{\{R_{i\tau} \leq R_{it}\}} \Big)
  \Big]^{-1} \,,\qquad i=1,\ldots,p;t=1,\ldots,n\,.
\end{equation*}
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.550]{EigenRatioSP500_ranks_log_50_shown.pdf}
  \caption{The logarithms of the ratios $\lambda_{(i+1)} / \lambda_{(i)}$ for the S\&P 500 series after rank transform.
We also show the 1, 50 and 99\% quantiles (bottom, middle, top lines, respectively) of the variables
$\log((\Gamma_i/ \Gamma_{i+1})^{2})$. }
  \label{fig:EigenRatio}
\end{figure}
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.550]{EigenRatioSP500_log_50_shown.pdf}
  \caption{The logarithms of the ratios $\lambda_{(i+1)} / \lambda_{(i)}$ for the original (non-rank transformed) S\&P 500 log-return data.
We also show the 1, 50 and 99\% quantiles (bottom, middle, top lines, respectively) of the variables
$\log((\Gamma_i/ \Gamma_{i+1})^{2/2.3})$; see also Figure~\ref{fig:EigenRatio} for comparison.}\label{fig:22}
\end{figure}

@@

\subsection{A point process convergence result}
Assume that the conditions at the beginning of Appendix \ref{appendix:A} hold.
Consider a sequence of iid copies $(S_{n}^{(t)})_{t=1,2,\ldots}$
of $S_n$ and the sequence of point processes
\begin{equation*}
N_n= \sum_{t=1}^{p} \vep_{a_{np}^{-1} S_{n}^{(t)}}, \quad n=1,2,\ldots\,,
\end{equation*}
for an integer sequence $p=p_n\to\infty$. We assume that the state space of the
point processes $N_n$ is $\overline{\R}_0=[\R\cup\{\pm \infty\}]\backslash \{0\}$.
\begin{lemma}\label{lem:ppr}
Assume $\alpha \in (0,2)$ and the
conditions of  Appendix~\ref{appendix:A}
on the iid sequence $(Z_t)$ and the normalizing sequence $(a_n)$. Then the limit relation
$N_n\cid N$ holds in the space of point measures on $\overline{\R}_0$
equipped with the vague topology (see
\cite{resnick:1987,resnick:2007})
for a  Poisson random measure $N$ with state space $\overline{\R}_0$ and intensity measure $\mu_\alpha(dx)=\alpha |x|^{-\alpha-1} (p_+ \1_{\{x>0\}}+ p_- \1_{\{x<0\}}) dx$.
\end{lemma}
\begin{proof}
According to Resnick \cite{resnick:1987}, Proposition 3.21, we need to
show that
$p\, \P(a_{np}^{-1}S_n\in \cdot)\civ \mu_\alpha
$,
where $\civ$ denotes vague convergence of Radon measures on  $\overline{\R}_0$.
Observe that we have $a_{np}/a_n\to\infty$ as $\nto$. This fact and
$\alpha\in (0,2)$ allow one to apply
Theorem~\ref{thm:nagaev}:
\begin{equation*}
\dfrac{\P( S_n >x a_{np})}{n\,\P(|Z|>  a_{np})}\to p_+ x^{-\alpha}
\quad \mbox{and}\quad \dfrac{\P( S_n \le -x a_{np})}{n\,\P(|Z|>
  a_{np})}\to p_-\, x^{-\alpha}\,,\quad x>0\,.
\end{equation*}
On the other hand, $n\,\P(|Z|>  a_{np})\sim p^{-1}$ as $\nto$.
This proves the lemma.
\end{proof}

>>>>

To illustrate the theory we consider a simple moving average example
taken from Davis et al.~\cite{davis:mikosch:pfaffel:2015}.
\begin{example} \rm
\begin{figure}[htb!]
  \centering
  \subfigure[iid data]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps_iid.pdf}
  }
  \subfigure[data from model \eqref{ex:4.6}]{
    \includegraphics[scale=0.4]{lambda_1_2_Gaps.pdf}
  }
  \caption{Distribution function of $(\lambda_{(1)} - \lambda_{(2)})/\lambda_{(1)}$ for iid data (left) and
    data generated from the model \eqref{ex:4.6} (right). In each graph we compare the empirical \ds\ \fct\
(dotted line, based on
1000 simulations of $200 \times 1000$ matrices with $Z$-\ds\ \eqref{eq:distrsim}) with the theoretical curve (solid line).}
  \label{fig:ProbMass}
\end{figure}
Assume that $\alpha\in (0,2)$ and
\begin{equation}\label{ex:4.6}
X_{it}= Z_{it}+ Z_{i,t-1}-2 (Z_{i-1,t}- Z_{i-1,t-1})\,,\quad i,t\in\Z\,.
\end{equation}
In this case, the non-zero entries of $\H$ are
\begin{equation*}
h_{00}=1, h_{01}=1,h_{10}=-2 \quad \mbox{ and }\quad h_{11}=2.
\end{equation*}
Hence $\M=\H\H'$ has the positive eigenvalues
$v_1=8$ and $v_2=2$. The limit point process in \eqref{eq:Davispp} is
\begin{equation*}
N=\sum_{i=1}^\infty \vep_{8\Gamma_i^{-2/\alpha}}+ \sum_{i=1}^\infty \vep_{2\Gamma_i^{-2/\alpha}}\,,
\end{equation*}
so that
\begin{equation*}
a_{np}^{-2}\,\big(\la_{(1)},\la_{(2)}\big) \cid
\big(8\Gamma_{1}^{-2/\alpha}, 2\Gamma_1^{-2/\alpha}\vee 8
\Gamma_2^{-2/\alpha}\big)\,.
\end{equation*}
Using the fact that $U=\Gamma_1/\Gamma_2$ has a uniform \ds\ on $(0,1)$ we calculate
\begin{equation*}
\P(2\Gamma_1^{-2/\alpha}>8\Gamma_2^{-2/\alpha})= \P(\Gamma_1/\Gamma_2<2^{-\alpha})= 2^{-\alpha} \in (1/4,1).
\end{equation*}
In particular, we have for the normalized spectral gap
\begin{equation*}
a_{np}^{-2} \big(\la_{(1)}-\la_{(2)}\big)\cid
6 \,\Gamma_1^{-2/\alpha} \I_{\{\Gamma_1 4^{\alpha/2 }<\Gamma_2\}}+
8\,
\big(\Gamma_1^{-2/\alpha}-\Gamma_2^{-2/\alpha}\big)\I_{\{\Gamma_14^{\alpha/2
  }> \Gamma_2\}}
\end{equation*}
and for the self-normalized spectral gap (see also Example~\ref{exam:spectralgap} for a detailed analysis)
\begin{equation*}
\begin{split}
\dfrac{\la_{(1)}-\la_{(2)}}{\la_{(1)}}&\cid
\dfrac{6}{8} \, \I_{\{\Gamma_1 2^{\alpha }<\Gamma_2\}}+
\big(1-(\Gamma_1/\Gamma_2)^{2/\alpha}\big)\I_{\{\Gamma_12^{\alpha
  }> \Gamma_2\}}\\
& =\dfrac{3}{4}\,  \I_{\{U 2^{\alpha }<1\}}+
\big(1-U^{2/\alpha}\big)\I_{\{U2^{\alpha
  }> 1\}}=Y\,.
\end{split}
\end{equation*}
The limit  distribution of the spectral gap
has an atom at $3/4$ with probability $2^{-\alpha}$, i.e.,~$\P(Y=3/4)=2^{-\alpha}$, and
\begin{equation*}
\P(Y\le x)= 1- (1-x)^{\alpha/2},\qquad x\in (0,3/4).
\end{equation*}
In the iid case the limit distribution of the self-normalized spectral gap has distribution function $F(x) = 1- (1-x)^{\alpha/2}$ for $x\in [0,1]$. This means that the atom disappears if the entries are iid. Figure~\ref{fig:ProbMass} compares the distribution function of $Y$
with $F$ for $\alpha=0.6$; the atom at $3/4$ is clearly visible.

@@

8\,\Gamma_1^{-2/\alpha},\frac{1}{4}\,\I_{\{U<2^{-\alpha}\}}+U^{2/\alpha}\,\I_{\{U\ge

@@

uniform \ds , $2^{-\alpha}\, \I_{\{U<2^{-\alpha}\}}+U\, \I_{\{U\ge 2^{-\alpha} \}}$.

@@

\frac{v_2}{v_1}\,\I_{\{U<(v_2/v_1)^{\alpha/2}\}}+U^{2/\alpha}\,\I_{\{U\ge (v_2/v_1)^{\alpha/2}\}}\,.

@@

\dfrac{\la_{(1)}-\la_{(2)}}{\la_{(1)}} \cid \dfrac{v_1-v_2}{v_1}\,\I_{\{U<(v_2/v_1)^{\alpha/2}\}}+(1-U^{2/\alpha})\,\I_{\{U\ge (v_2/v_1)^{\alpha/2}\}}\,.

@@

\subsection{S\&P 500 data}\label{sec:sp500}
We conduct a short analysis of the largest eigenvalues of the univariate log-return time series which compose
the S\&P 500 stock index; see Section~\ref{subsec:1.2} for a description of the data.
Although there is strong empirical evidence that these univariate series have power-law tails
(see Figure~\ref{fig:SP500_tail_indices}) we  do not expect that they
have the same tail index. One way to proceed would be to ignore this fact because the tail indices are
in a close range and the differences are due to large sampling errors for estimating such quantities.  One could also collect \ts\ with similar
tail indices in the same group.
In this case, the dimension $p$ decreases. This grouping would be a rather arbitrary classification method.
We have chosen a third way: to use rank transforms.
This approach has its merits because it aims at standardizing the tails but it also has a major disadvantage: one destroys
the covariance structure underlying the data.
\par
Given a $p\times n$ matrix $(R_{it})_{i = 1, \cdots, p;t=1,\cdots, n}$,
we construct a matrix $\X$ via the rank transforms
\begin{equation*}
  X_{it} = -\Big[
    \log \Big(\frac{1}{n+1} \sum_{\tau=1}^n \I_{\{R_{i\tau} \leq R_{it}\}} \Big)
  \Big]^{-1} \,,\qquad i=1,\ldots,p;t=1,\ldots,n\,.
\end{equation*}
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.550]{EigenRatioSP500_ranks_log_50_shown.pdf}
  \caption{The logarithms of the ratios $\lambda_{(i+1)} / \lambda_{(i)}$ for the S\&P 500 series after rank transform.
We also show the 1, 50 and 99\% quantiles (bottom, middle, top lines, respectively) of the variables
$\log((\Gamma_i/ \Gamma_{i+1})^{2})$. }
  \label{fig:EigenRatio}
\end{figure}
\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.550]{EigenRatioSP500_log_50_shown.pdf}
  \caption{The logarithms of the ratios $\lambda_{(i+1)} / \lambda_{(i)}$ for the original (non-rank transformed) S\&P 500 log-return data.
We also show the 1, 50 and 99\% quantiles (bottom, middle, top lines, respectively) of the variables
$\log((\Gamma_i/ \Gamma_{i+1})^{2/2.3})$; see also Figure~\ref{fig:EigenRatio} for comparison.}\label{fig:22}
\end{figure}

@@

\subsection{A point process convergence result}
Assume that the conditions at the beginning of Appendix \ref{appendix:A} hold.
Consider a sequence of iid copies $(S_{n}^{(t)})_{t=1,2,\ldots}$
of $S_n$ and the sequence of point processes
\begin{equation*}
N_n= \sum_{t=1}^{p} \vep_{a_{np}^{-1} S_{n}^{(t)}}, \quad n=1,2,\ldots\,,
\end{equation*}
for an integer sequence $p=p_n\to\infty$. We assume that the state space of the
point processes $N_n$ is $\overline{\R}_0=[\R\cup\{\pm \infty\}]\backslash \{0\}$.
\begin{lemma}\label{lem:ppr}
Assume $\alpha \in (0,2)$ and the
conditions of  Appendix~\ref{appendix:A}
on the iid sequence $(Z_t)$ and the normalizing sequence $(a_n)$. Then the limit relation
$N_n\cid N$ holds in the space of point measures on $\overline{\R}_0$
equipped with the vague topology (see
\cite{resnick:1987,resnick:2007})
for a  Poisson random measure $N$ with state space $\overline{\R}_0$ and intensity measure $\mu_\alpha(dx)=\alpha |x|^{-\alpha-1} (p_+ \I_{\{x>0\}}+ p_- \I_{\{x<0\}}) dx$.
\end{lemma}
\begin{proof}
According to Resnick \cite{resnick:1987}, Proposition 3.21, we need to
show that
$p\, \P(a_{np}^{-1}S_n\in \cdot)\civ \mu_\alpha
$,
where $\civ$ denotes vague convergence of Radon measures on  $\overline{\R}_0$.
Observe that we have $a_{np}/a_n\to\infty$ as $\nto$. This fact and
$\alpha\in (0,2)$ allow one to apply
Theorem~\ref{thm:nagaev}:
\begin{equation*}
\dfrac{\P( S_n >x a_{np})}{n\,\P(|Z|>  a_{np})}\to p_+ x^{-\alpha}
\quad \mbox{and}\quad \dfrac{\P( S_n \le -x a_{np})}{n\,\P(|Z|>
  a_{np})}\to p_-\, x^{-\alpha}\,,\quad x>0\,.
\end{equation*}
On the other hand, $n\,\P(|Z|>  a_{np})\sim p^{-1}$ as $\nto$.
This proves the lemma.
\end{proof}
